---
title: "Finding and playing with peaks in RMINC"
output: html_document
author: Jason Lerch
date: 2018-02-09
---

So, peaks. When producing a statistical map, it's good to get a report of the peaks (i.e. most significant findings). RMINC has had this support for a while now, though it has remained somewhat hidden. Here's a bit of an intro, then.

I will walk through the example we used from the Mouse Imaging Summer School in 2017, which is data from this paper:

de Guzman AE, Gazdzinski LM, Alsop RJ, Stewart JM, Jaffray DA, Wong CS, Nieman BJ. Treatment age, dose and sex determine neuroanatomical outcome in irradiated juvenile mice. Radiat Res. 2015 May;183(5):541–9.

To keep it simple, however, I'll only look at sex differences in that dataset for now.

Let's start - load the libraries and read in the csv file that describes the data.

```{r}
suppressMessages(library(RMINC))
gf <- read.csv("/hpf/largeprojects/MICe/jason/MISS2017/intro-stats/fixed_datatable_IRdose.csv")
```

And run a linear model relating the Jacobian determinants to sex and radiation dose. I'll use the segmentations file as a mask; it'll be needed later on anyway.

```{r}
labelFile <- "/hpf/largeprojects/MICe/jason/MISS2017/intro-stats/atlas-registration/pipeline-18-08-2017-at-07-08-48_processed/nlin-3/voted.mnc"
vs <- mincLm(Jacobfile_scaled0.2 ~ Sex + Dose, gf, mask = labelFile)
```

Some more data preparation: read in the background anatomy file ...

```{r}
anatFile <- "/projects/moush/lbernas/Irradiation_behaviour_project/fixed_build_masked_23mar13_nlin/nlin-3.mnc"
anatVol <- mincArray(mincGetVolume(anatFile))
```

... and show the results at a somewhat arbitrary threshold.

```{r}
mincPlotSliceSeries(anatVol, mincArray(vs, "tvalue-SexM"), anatLow=10, anatHigh=15, low=2, high=6, symmetric = T,
                    begin=50, end=-50)
```


At this point we've run a linear model and visually assessed the results. Now we can locate the peak findings, using the `mincFindPeaks` command.

```{r}
# set the path; should usually be unnecessary, but my remote Rstudio setup appears to not load the correct environment variables 
Sys.setenv(PATH = paste("/axiom2/projects/software/arch/linux-xenial-xerus/minc-toolkit/1.9.15/bin", Sys.getenv("PATH"), sep=":"))
# and the actual command
peaks <- mincFindPeaks(vs, "tvalue-SexM", minDistance = 1, threshold = 4)
```

`mincFindPeaks` uses the `find_peaks` command from the MINC toolkit under the hood. You pass in the output of one of the RMINC modelling commands (mincLm in this case, but can be anything), along with the column from that model you want to get peaks from. You can then set the minimum distance between peaks (in mm) - i.e. how far apart do two statistical peaks have to be to be included? - as well as the threshold to be considered a peak. Optionally thresholds can be different for positive and negative peaks; as always, see `?mincFindPeaks` for more detail.

This is what we have at this point:

```{r, eval = FALSE}
peaks
```
```{r, results = "asis", echo = FALSE}
DT::datatable(peaks, options = list(pageLength = 5))
```

There are 7 columns; the first three give the coordinates in voxel space, the next three the coordinates in world space, and then the peak value itself. There's a further helpful command that labels the peaks with the atlas location within which the peak is located:

```{r}
defsFile <- "/hpf/largeprojects/MICe/tools/atlases/Dorr_2008_Steadman_2013_Ullmann_2013_Richards_2011_Qiu_2016_Egan_2015_40micron/mappings/DSURQE_40micron_R_mapping.csv"
peaks <- mincLabelPeaks(peaks, labelFile, defsFile)
```

This adds an extra column containing the name of the label for each peak:

```{r, eval = FALSE}
peaks
```
```{r, results = "asis", echo = FALSE}
DT::datatable(peaks, options = list(pageLength = 5))
```

At this point you have all the info about the most significant peaks in the dataset. There is one additional useful command, a shortcut to plot each peak. Plotting the two most positive and the two most negative peaks, for example:

```{r}
opar <- par(mfrow=c(2,2), mar=c(0,0,0,0))
nP <- nrow(peaks)
for (i in c(1:2, nP:(nP-1))) {
  mincPlotPeak(peaks[i,], anatVol, mincArray(vs, "tvalue-SexM"), anatLow=10, anatHigh=15, low=2, high=6, symmetric=T)
}
par(opar)
```

There is an additional argument to `mincPlotPeak` - a function to create a plot of the peak. This is illustrated below:

```{r}
# load ggplot2 for plotting
library(ggplot2)
# the plotting function; needs to take a single argument, the peak
p <- function(peak) {
  # read in the data for that particular peak
  gf$voxel <- mincGetWorldVoxel(gf$Jacobfile_scaled0.2, peak["x"], peak["y"], peak["z"])
  # and create a box-plot; also read in info from the peak for a meaningful title
  ggplot(gf) + aes(x=Sex, y=exp(voxel)) + geom_boxplot() + 
    ggtitle(paste("Peak:", peak["label"]), subtitle = paste("T statistic:", peak["value"]))
}
# and plot the peak with its plot
mincPlotPeak(peaks[1,], anatVol, mincArray(vs, "tvalue-SexM"), anatLow=10, anatHigh=15, low=2, high=6, 
             symmetric=T, plotFunction = p)
```

And that's it. A quick survey for how to extract peaks, view them in a table, and create figures of the most significant findings.
